Covering Your Basis: Model-Agnostic Inference of LIGO-Virgos Binary Black Hole Spin Distribution

### 2022-08-30 src/tex/ms.tex
setup zenodo integration

\author{Lexi Vives}
\affiliation{Institute  for  Fundamental  Science, Department of Physics, University of Oregon, Eugene, OR 97403, USA}

@@

Now that gravitational wave observations of compact binary coalescence is a regular occurrence, producing catalogs of nearly 
100 such detections, we now have access to a complete understanding of the population of compact binaries. In the coming fourth 
observing run of the LIGO-Virgo-KAGRA collaboration, the catalog size is expected to increase by a factor of 5. The extensive 
physics governing the formation and evolution of compact objects that end up merging is imprinted on the properties of the population 
of CBCs, allowing us to learn about these not well understood ideas through looking at the distribution of observed GWs. In particular 
the mass distribution of binary black holes has been shown to be very useful in measuring cosmological parameters, constraining modified 
gravitational wave propagation, constraining a running Planck mass, looking of evidence of ultralight bosons through superradiance, 
putting constraints on uncertain nuclear reaction rates important for stellar evolution, and even looking for primordial black holes. 
Through a better understanding of the mass, spin, and redshift distribution of compact binaries we can extend this be a useful tool 
in all the different physical applications mentioned above. In this paper we highlight how the use of basis-splines can be a useful 
model-agnostic modeling approach to the astrophysical distributions of compact binaries. We show that with a particular choice of basis, 
we can efficiently model the spin and/or mass distribution of compact binaries in GWTC-3 with spline regression in hierarchical 
Bayesian inference. We show our results are similar to those produced with the more simpler and motivated parametric models, 
while being capable of finding more subtle features should they appear in future catalogs. This method can be extended to higher-dimensional 
basis-splines to model correlations between different distributions which will be necessary to confidently disentangle differently 
competing formation channels should they exist.

@@

\subsection{Mass Spectrum Inference with GWTC-3} \label{sec:GWTC3_inferece}
\section{Conclusions}\label{sec:conclusion}
\section{Acknowledgements}\label{sec:acknowledments}
This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center 
(\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. 
This work benefited from access to the University of Oregon high performance computer, Talapas.  This material is based upon work supported 
in part by the National Science Foundation under Grant PHY-1807046 and work supported by NSF’s LIGO Laboratory which is a major facility 
fully funded by the National Science Foundation.
\software{
\textsc{Astropy}~\citep{2018AJ....156..123A},
\textsc{NumPy}~\citep{harris2020array},
\textsc{Matplotlib}~\citep{Hunter:2007},
\textsc{bilby}~\citep{Ashton_2019},
\textsc{GWPopulation}~\citep{Talbot_2019}
}
\bibliography{references}{}
\bibliographystyle{aasjournal}

>>>>

@@

Now that gravitational wave observations of compact binary coalescence is a regular occurrence, producing catalogs of nearly 
100 such detections, we are beginning to have access to a more complete understanding of the population of compact binaries. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA collaboration, the catalog size is expected to increase by up to a factor of 5. 
Looking at population properties of compact objects lets us probe the complex physics common to all of our observations that governs the 
formation and evolution of merging compact objects. The history of these dense objects is imprinted on the properties of the population 
of CBCs, allowing us to learn about these not well understood ideas through careful evaluation of the enture distribution of observed GWs. 
The mass distribution of binary black holes has been used to measure cosmological parameters, constrain modified gravitational wave propagation, 
constrain a non-constant Planck mass, look for evidence of ultralight bosons through superradiance, put constraints on the uncertain nuclear reaction 
rates important for stellar evolution, look for primordial black holes, to constrain the neutron star equation of state, 
and to understand the fromation channels/environments of compact binary coalescences (CBCs). Through a better understanding of the mass, spin, 
and redshift distribution of compact binaries that will come with the increased catalog size, one can probe these dufferent physical phenoma with 
greater fidelity.

With the latest LIGO-Virgo-KAGRA (LVK) catalog of CBCs (GWTC-3), new understanding of the population has come into light. One of the main findings was that 
the mass distribution exhibits substructure with peaks rather than being perfectly described by a smooth power law distribution. The evidence for this was 
found with mutliple different flexible or non-parametric models that aim to be data-oriented with little bias on the function form. While these models are 
useful for looking for unexpected features in the data, or finding where our simpler models fail, they can be harder to interpret and are not based on specific
astrophysical motivation. These types of models will become more important as the size of our catalogs increase and we enter a data-dominated regime rather than
prior/model dominated.

The spin distribution of binary black holes (BBHs) has also been of controversial interest in the literature recently, with different parameteric approaches finding 
rather different conclusions on the distribution of spins of BBHs which has significant impact on the inferences about relative contribution of dynamically formed
binaries versus ones formed in the field. This is another use case for flexible data-driven models to alleivate the discrepencies among parametric approaches. 
In particular there has been little work on flexible model development for the spin distribution of BBHs, which makes sense given the relatively smaller catalog 
sizes and the large uncertainty in individual black hole spin measurements with GWs.

In this paper we highlight how the use of basis-splines can be a useful model-agnostic modeling approach to the astrophysical distributions of compact 
binaries. We show that with a particular choice of basis, one can efficiently model the spin, mass and redshift distributions of compact binaries in GWTC-3 with
hierarchical Bayesian inference. We discuss our results in the context of current literature on compact object populations and illustrate how our results are similar
to those produced with the more simpler and astrophysically motivated parametric models. This data-drien approach will be capable of finding more subtle features, that 
may not be expected, should they appear in future catalogs. Additionally this method can be extended to higher-dimensional basis-splines in order to model 
joint population disributions over multiple surce parameters, which will be important to confidently disentangle different competing formation channels 
should they exist. In Section II we go over the background of basis splines and model construction. In Section III we present results using our basis spline model 
on the mass and spin distributions of BBHs in GWTC-3, followed by discussing the results and astrophysical implications in Section IV and concluding in Section V.

@@

\subsection{Mass Spectrum Inference with GWTC-3} \label{sec:GWTC3_inferece}
\section{Conclusions}\label{sec:conclusion}
\section{Acknowledgements}\label{sec:acknowledments}
This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center 
(\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. 
This work benefited from access to the University of Oregon high performance computer, Talapas.  This material is based upon work supported 
in part by the National Science Foundation under Grant PHY-1807046 and work supported by NSF’s LIGO Laboratory which is a major facility 
fully funded by the National Science Foundation.
\software{
\textsc{Astropy}~\citep{2018AJ....156..123A},
\textsc{NumPy}~\citep{harris2020array},
\textsc{Matplotlib}~\citep{Hunter:2007},
\textsc{bilby}~\citep{Ashton_2019},
\textsc{GWPopulation}~\citep{Talbot_2019}
}
\bibliography{bib}{}
\bibliographystyle{aasjournal}
